Leadership across cultures and contexts: an exploratory analysis

### 2020-05-11 leadership_across_cultures_contexts.Rmd
Revising discussion section on status.

Leadership can evolve when the benefits exceed the costs, and the same for followership. Our data did not allow us to compare benefits and costs. Instead, we compared levels of evidence for benefits vs. costs for each leader benefit/cost dimension and each follower benefit/cost dimension. Specifically, we fit a mixed effects logistic regression model of evidence (0/1) as a function of each dimension (e.g., Mating, Food), a term indicating if the evidence was for a benefit or cost, and their interaction, with random intercepts for author nested within culture. We fit separate models for leaders and followers. These two models allowed us to estimate the relative evidence for, e.g., leader mating benefits vs. leader mating costs, and follower mating benefits vs. follower mating costs, etc. For leaders, the odds ratio that a text record would provide evidence for a benefit vs. a cost (averaging across all benefit/cost dimensions), was $OR=`r signif(leader_bc_ratio$benefit_cost_OR, 2)`$, and for followers, was $OR=`r signif(follower_bc_ratio$benefit_cost_OR, 2)`$. Hence for both leaders and followers, there was more evidence of benefits than costs.

@@

We found strong evidence that several dimensions of leadership were universal across cultures, and across group contexts within cultures. Leaders function to resolve conflict, organize cooperation, and provide counsel and direction in over 70% of cultures, and are seen as high status, knowledgeable or intelligent, and experienced or accomplished in about 80% or more of cultures (see Figure \@ref(fig:varsupport)). Leaders benefited materially, reproductively, or socially in 50% or more of cultures (see Figure \@ref(fig:varsupportcb)). The composite feature measures mirrored these results, finding evidence in 85% of cultures or more, that leaders had social, reproductive, or material success, exhibited a number of competencies, organized and strategized group cooperation, and were prosocial. See Figure \@ref(fig:featuressupportplot).

@@

The leadership "features" we derived by clustering leadership dimensions were represented across most cultures. Leaders can be conceptualized as performing three broad functions: providing prosocial goods and services, strategizing, and organizing collective actions. Regarding the traits characteristic of leadership, leaders can be conceptualized as possessing some combination of individual capacities, many of which are necessarily prosocial, conforming to particular cultural norms, and receiving various benefits in exchange for their functions. See Figure \@ref(fig:featuressupportplot).

@@

Most potentially universal dimensions discovered here appear in a model developed by @hagen_leadership_2019, in which knowledgeable and intelligent individuals provide computational (cognitive) services to group members in exchange for a variety of fitness benefits, and these services often involve conflict resolution and other social functions [see also @garfield_evolutionary_2019]. The computational services model emphasizes mating benefits, however, which appear in `r matingbenefit`% of cultures, below our arbitrary 60% threshold for "universality" (see Figure \@ref(fig:varsupportcb)). But there was evidence for the "Social material success" feature, which included Polygyny, High quality spouse, and Many children, in more cultures (`r reprosuccess`%) than any other feature (see Figure \@ref(fig:featuressupportplot)). Interestingly, the computational model highlights women's family leadership role, and the estimated marginal mean for Social, reproductive, and material success was higher for females leaders than male leaders (see Figure \@ref(fig:heatmapfeaturesex)), with the caveat that there were few text records on female leaders in our sample.

@@

The clearest pattern to emerge from analyses of systematic variation is that, not too surprisingly, many dimensions of leadership are context-specific: aggressiveness and bravery were associated with military leadership, supernatural qualities and ritual functions were associated with religious leadership, organizing cooperation and being experience or accomplished were associated with economic-group leadership, and providing counsel and older age were associated with kin-group leadership. See Figure \@ref(fig:heatmapgroups).

@@

We therefore coded all benefits and costs of both leadership and followership to better understand which benefits and costs are most commonly associated with leading and following. Leaders were widely reported to receive material benefits, social status, and mating benefits but loss of material resources and social status were also widely reported (see Figure \@ref(fig:varsupportcb)).

@@

All evolutionary models of leadership propose that the fitness benefits of leading outweigh the costs. The relative abundance of evidence for leader and follower benefits vs. costs ($OR=`r signif(leader_bc_ratio$benefit_cost_OR, 2)`$ and $OR=`r signif(follower_bc_ratio$benefit_cost_OR, 2)`$, resp.), while not providing evidence of net fitness benefits, is nevertheless encouraging, especially since leaders and followers both had more evidence of mating and territory benefits than costs, which are key contributors to fitness, and leaders also had more evidence for inclusive fitness, material and social benefits, relative to costs. See Figure \@ref(fig:benefitcostratio).

@@

Relationships between leadership and high status, including causality and dependence, are puzzling and currently the focus of much work in evolutionary leadership studies [e.g., @van_vugt_dual_2019; @smith_leadership_2020; @cheng_why_2020; @cheng_dominance_2019; @von_rueden_dynamics_2019]. Drawing on a comparative-evolutionary perspective, incorporating findings from nonhuman social species, @van_vugt_dual_2019 emphasize that although leadership and high social status do indeed often covary, leaders are not necessarily high-status individuals, exemplified when low-ranking individuals in nonhuman species emerge as leaders of group movement or when younger and lower status men and women initiate cooperative activities (such as raids in small-scale warfare or social-political movements). @cheng_dominance_2019 and @cheng_why_2020, emphasize, however, that social hierarchies form principally in response to inter-individual variation in status, which may be gained via distinct pathways. Thus, according to @cheng_why_2020, although all leaders are not necessarily high status individuals, especially across nonhuman species, among humans high social status may most often be a necessary but insufficient condition for achieving and maintaining positions of leadership. In our view, social context is critical to these discussions [@wiessner_power_2010], which is a point often mentioned but under-developed and under-investigated empirically. Processes of leadership and followership emerge across levels of social organization and individuals simultaneously maintain multiple social statuses. The "low status" male who initiates an inter-group raid on a neighboring group may not be widely respected within his community nor be a candidate for community leadership, however, he might be of exceptionally high status among his peers and high ranking within what @garfield_review_2019 [p. 71] called the "intragenerational social hierarchy."

Our results offer some insight into relationships between leadership and social status across cultures and contexts. As mentioned, high status is the most common quality of leaders (Figure \@ref(fig:varsupport)) and the second most common benefit of leadership and the loss of status the most frequent cost (Figure \@ref(fig:varsupportcb)). 
<!-- mentioned in this sample of the ethnographic record which includes discussions of chiefs, work group leaders, shamans, and age-sets, from a culturally diverse sample including foraging groups and kingdoms.  -->
This lends support to the notion of high social status often being a necessary condition of leader emergence [@cheng_why_2020]. In predicting the ethnographic texts which described leaders as high status, leaders of kin groups and leaders among hunter-gatherers and pastoralists were identified as more likely to be high status relative to some other group and subsistence types. Although discussions of female leaders were rare, the presence of female leaders were also associated with high social status (see Figure \@ref(fig:groupstatus)). Interestingly, political leaders were less likely to be described as high status, which could be related to institutionalized and overt or ascribed social distinctions. Nonetheless, these findings question the informed speculation by @smith_leadership_2020 that, "leadership may therefore be more strongly correlated with high status in large, complex organizations, such as in corporations and governments (p. 2)." However, our text analytic results, which relied on the semantic context of text records to predict the presence of high status leaders generally identified terms related to status and prestige (i.e., *respect*, *prestige*, *honor*) and possibly dominance or proscribed hierarchy (i.e., *rank*, *position*), as stronger predictors of high status leader than terms associated with leadership per se (i.e., *influence*, *authority*, *headman*), which lends some support to the distinction between status and leadership highlighted by @smith_leadership_2020 (though we caution from drawing strong interpretations from these text-analytic results, see Figure S\@ref(fig:highstatusplot)). Overall, these results suggest associations between leadership and high status are ubiquitous across human societies, including among kin groups and among hunter-gatherers and that while effective leaders across a wide range of social contexts are the recipients of increased social status for their valuable services, ineffective leadership is associated with reduced social status.

@@

<!-- DISCUSSES BOTH MANAGEMENT FEATURES -->
In the absence of state-institutions local leaders are often tasked with managing various economic systems and providing social services. Within the domain of leader management, there are likely to be two primary features of human leader functions: strategizing and organizing. Our results correspond very closedly to what @van_vugt_cognitive_2007 define as *strategic* versus *coordinating* leadership.

Leaders, possibly due to their special capacities, are able to devise and impliment strategic solutions to group problems including resolving conflicts within the group (the most widely supported function in our sample) and providing counsel (also widely supported), which again are functions that are commonly noted in the anthropological and psychological leadership literature [@wiessner_risk_1982; @wiessner_collective_2019; @boehm_hierarchy_1999; @knauft_violence_1991; @de_cremer_intergroup_2002], and concerning conflict resolution specifically, widespread among nonhuman animals as well [@smith_leadership_2016; @smith_obstacles_2018; @wilson_sociobiology:_1980]. Among the Mbuti, hunter-gatherers in the Congo Basin known for strong cultural values of personal autonomy and egalitarianism, @fo04-001 [p. 185] describes the nature of conflict resolution processes in the absence of strong authority:

@@

The feature of leader organizatonal functions aligns with the many evolutionary theories of leadership that emphasize the role of leaders in organizing collective actions [e.g., @hooper_theory_2010; @van_vugt_cognitive_2007; @glowacki_formation_2016; for review, see @garfield_evolutionary_2019]. In the ethnographic record such functions includes controlling and facilitating economic productivity and ensuring collective actions do not overly conflict with individual or household-level goals, such as among the Tukano of the Brazilian Amazon, where @sq19-018 [p. 127] notes:

>>>>

Leadership can evolve when the benefits exceed the costs, and the same for followership. Our data did not allow us to compare benefits and costs. Instead, we compared levels of evidence for benefits vs. costs for each leader benefit/cost dimension and each follower benefit/cost dimension. Specifically, we fit a mixed effects logistic regression model of evidence (0/1) as a function of each dimension (e.g., *Mating*, *Food*), a term indicating if the evidence was for a benefit or cost, and their interaction, with random intercepts for author nested within culture. We fit separate models for leaders and followers. These two models allowed us to estimate the relative evidence for, e.g., leader mating benefits vs. leader mating costs, and follower mating benefits vs. follower mating costs, etc. For leaders, the odds ratio that a text record would provide evidence for a benefit vs. a cost (averaging across all benefit/cost dimensions), was $OR=`r signif(leader_bc_ratio$benefit_cost_OR, 2)`$, and for followers, was $OR=`r signif(follower_bc_ratio$benefit_cost_OR, 2)`$. Hence for both leaders and followers, there was more evidence of benefits than costs.

@@

We found strong evidence that several dimensions of leadership were universal across cultures, and across group contexts within cultures. Leaders function to resolve conflict, organize cooperation, and provide counsel or direction in over 70% of cultures, and are seen as high status, knowledgeable or intelligent, and experienced or accomplished in about 80% or more of cultures (see Figure \@ref(fig:varsupport)). Leaders benefited materially, reproductively, or socially in 50% or more of cultures (see Figure \@ref(fig:varsupportcb)). The composite feature measures mirrored these results, finding evidence in 85% of cultures or more, that leaders had social, reproductive, or material success, exhibited a number of competencies, organized and strategized group cooperation, and were prosocial. See Figure \@ref(fig:featuressupportplot).

@@

The leadership features we derived by clustering leadership dimensions were represented across most cultures. Leaders can be conceptualized as performing three broad functions: providing prosocial goods and services, strategizing, and organizing collective actions. Regarding the traits characteristic of leadership, leaders can be conceptualized as possessing some combination of individual capacities, many of which are necessarily prosocial, conforming to particular cultural norms, and receiving various benefits in exchange for their functions. See Figure \@ref(fig:featuressupportplot).

@@

Most potentially universal dimensions discovered here appear in a model developed by @hagen_leadership_2019, in which knowledgeable and intelligent individuals provide computational (cognitive) services to group members in exchange for a variety of fitness benefits, and these services often involve conflict resolution and other social functions [see also @garfield_evolutionary_2019]. The computational services model emphasizes mating benefits, however, which appear in `r matingbenefit`% of cultures, below our arbitrary 60% threshold for "universality" (see Figure \@ref(fig:varsupportcb)). But there was evidence for the "Social, reproductive, and materials success" feature, which included leaders in polygynous unions, having a high quality spouse, and having many children, in more cultures (`r reprosuccess`%) than any other feature (see Figure \@ref(fig:featuressupportplot)). Interestingly, the computational model highlights women's family leadership role, and the estimated marginal mean for the *Social, reproductive, and materials success* feature was higher for females leaders than male leaders (see Figure \@ref(fig:heatmapfeaturesex)), with the caveat that there were few text records on female leaders in our sample.

@@

The clearest pattern to emerge from analyses of systematic variation is that, not too surprisingly, many dimensions of leadership are context-specific: aggressiveness and bravery were associated with military leadership, supernatural qualities and ritual functions were associated with religious leadership, organizing cooperation and being experience or accomplished were associated with economic-group leadership, and providing counsel and older age were associated with kin-group leadership. See Figure \@ref(fig:heatmapgroups). These results underscore recent trends in evoltuionary approaches to leadership which emphasize the context-dependency of leader emergence and evolutionary selection pressures [@smith_leadership_2020; @smith_leadership_2016; @hagen_leadership_2019].

@@

We therefore coded all benefits and costs of both leadership and followership to better understand which benefits and costs are most commonly associated with leading and following. Leaders were widely reported to receive material benefits, social status, and mating benefits but loss of material resources and social status were also widely reported, suggesting leadership to be a high risk-return strategy (see Figure \@ref(fig:varsupportcb)).

@@

All evolutionary models of leadership propose that the fitness benefits of leading outweigh the costs [e.g., @van_vugt_leadership_2008; @kantner_identifying_2010]. The relative abundance of evidence for leader and follower benefits vs. costs ($OR=`r signif(leader_bc_ratio$benefit_cost_OR, 2)`$ and $OR=`r signif(follower_bc_ratio$benefit_cost_OR, 2)`$, resp.), while not providing evidence of net fitness benefits, is nevertheless encouraging, especially since leaders and followers both had more evidence of mating and territory benefits than costs, which are key contributors to fitness, and leaders also had more evidence for inclusive fitness, material and social benefits, relative to costs. These results provide a framework for further work on evolutionarily relevent costs and benfits of leadership, and suggests leaders-follower dynamics should continue to be prioritized in evolutionary models of demography, group dynamics, social complexity and inequality, cooperation and punishment, and resource exchange [c.f., @bird_variability_2019; @ember_our_2018; @micheletti_modelling_2020; @smith_cultural_2020; @birch2017philosophy; @marlowe_more_2008; @henrich_markets_2010; @mace_co-evolution_1998; @mace_phylogenetic_2005; @mattison_evolution_2016] See Figure \@ref(fig:benefitcostratio).

@@

Relationships between leadership and high status, including causality and dependence, are puzzling and currently the focus of much work in evolutionary leadership studies [e.g., @van_vugt_dual_2019; @smith_leadership_2020; @cheng_why_2020; @cheng_dominance_2019; @von_rueden_dynamics_2019]. Drawing on a comparative-evolutionary perspective, @van_vugt_dual_2019 emphasize that although leadership and high social status do often covary, leaders are not always high-status individuals, exemplified when low-ranking individuals in nonhuman species lead group movement or when younger and lower status men and women initiate collective actions (as examples the authors point to raids in small-scale warfare and widespread social-political movements). @cheng_dominance_2019 and @cheng_why_2020, emphasize, however, that social hierarchies form principally in response to inter-individual variation in status, which may be gained via distinct pathways. Thus, according to @cheng_why_2020, although all leaders are not necessarily high status individuals, especially across nonhuman species, among humans high social status may most often be a necessary but insufficient condition for achieving and maintaining positions of leadership. In our view, social context is critical to these discussions [see @wiessner_power_2010]. Leadership and followership emerge across levels of social organization and individuals simultaneously maintain multiple social "statuses." The "low status" male who initiates an inter-group raid may not be widely respected nor a candidate for community leadership, however, he may be high status among his peers and within what @garfield_review_2019 [p. 71] described as the "intragenerational social hierarchy."

Our results offer some insight into relationships between leadership and social status across cultures and contexts. As mentioned, high status is the most common quality of leaders (Figure \@ref(fig:varsupport)) and the second most common benefit of leadership and the loss of status the most frequent cost (Figure \@ref(fig:varsupportcb)). 
<!-- mentioned in this sample of the ethnographic record which includes discussions of chiefs, work group leaders, shamans, and age-sets, from a culturally diverse sample including foraging groups and kingdoms.  -->
This lends support to the notion of high social status often being a necessary condition of leader emergence [@cheng_why_2020]. The dimensions of leadership predictive of high status leaders suggest the phenotype of such leaders is linked to specific competencies, including experience, decision-making, intelligence, and inter-personal skills, as well as older age and conformity to social norms, which does support claims of independence by @smith_leadership_2020 [see @von_rueden_multiple_2008]. High status leaders are able to direct benefits to their close relatives, likely in part to their greater material and social wealth, which implicates leadership dynamics as a driver of the evolution and emergence of inequality [c.f., @sanday_female_1981; @shenk_status_2016; @price_social_2010; @borgerhoff_mulder_pastoralism_2010; @mattison_evolution_2016; @bowles_emergence_2010]. Our results did not strongly implicate high status with particular functions, with the exception that high status leaders are unlikely to enforce sanctions on group members. See Figure \@ref(fig:highstatusplot)B.

Although evidence for high status leaders did not substantially vary by subsistence or group structure, this null result suggests relationships between leadership and social status are not specific to social complexity or substantially context-dependent and if anything leadership among lower levels of social organization and less politically complex societies (i.e., kin groups and hunter-gatherers) is more likely to be associated with high social status, relative to other groups (see \@ref(fig:groupstatus)A and \@ref(fig:groupstatus)B), which could be related to institutionalized and overt or ascribed social distinctions among more politically complex populations. Nonetheless, these findings question the informed speculation by @smith_leadership_2020 that, "leadership may therefore be more strongly correlated with high status in large, complex organizations, such as in corporations and governments (p. 2)." However, our text analytic results, which relied on the semantic context of text records to predict the presence of high status leaders generally identified terms related to status and prestige (i.e., *respect*, *prestige*, *honor*) and possibly dominance or proscribed hierarchy (i.e., *rank*, *position*), as stronger predictors of high status leader than terms associated with leadership per se (i.e., *influence*, *authority*, *headman*), which lends some support to the distinction between status and leadership highlighted by @smith_leadership_2020 (though we caution from drawing strong interpretations from these text-analytic results, see Figure S\@ref(fig:highstatusplot)). Overall, these results suggest associations between leadership and high status are ubiquitous across human societies, including among kin groups and among hunter-gatherers and that while effective leaders across a wide range of social contexts are the recipients of increased social status for their valuable services, ineffective leadership is associated with reduced social status.

Although discussions of female leaders were rare, the presence of female leaders were also associated with high social status (see Figure \@ref(fig:groupstatus)).

@@

<!-- DISCUSSES BOTH MANAGEMENT FEATURES -->
In the absence of state-institutions local leaders are often tasked with managing various economic systems and providing social services. Within the domain of leader management, there are likely to be two primary features of human leader functions: strategizing and organizing. Our results correspond very closely to what @van_vugt_cognitive_2007 define as *strategic* versus *coordinating* leadership.

Leaders, possibly due to their special capacities, are able to devise and implement strategic solutions to group problems including resolving conflicts within the group (the most widely supported function in our sample) and providing counsel (also widely supported), which again are functions that are commonly noted in the anthropological and psychological leadership literature [@wiessner_risk_1982; @wiessner_collective_2019; @boehm_hierarchy_1999; @knauft_violence_1991; @de_cremer_intergroup_2002], and concerning conflict resolution specifically, widespread among nonhuman animals as well [@smith_leadership_2016; @smith_obstacles_2018; @wilson_sociobiology:_1980]. Among the Mbuti, hunter-gatherers in the Congo Basin known for strong cultural values of personal autonomy and egalitarianism, @fo04-001 [p. 185] describes the nature of conflict resolution processes in the absence of strong authority:

@@

The feature of leader organizational functions aligns with the many evolutionary theories of leadership that emphasize the role of leaders in organizing collective actions [e.g., @hooper_theory_2010; @van_vugt_cognitive_2007; @glowacki_formation_2016; for review, see @garfield_evolutionary_2019]. In the ethnographic record such functions includes controlling and facilitating economic productivity and ensuring collective actions do not overly conflict with individual or household-level goals, such as among the Tukano of the Brazilian Amazon, where @sq19-018 [p. 127] notes:
